---
title: "Response Surface Models with Mplus"
description: |
  Generate, write, run, read, and plot latent response surface models with
  rsahelpers, tidySEM, and MplusAutomation.
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
vignette: >
  %\VignetteIndexEntry{Response Surface Models with Mplus}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: true
  warning: false
  message: false
---

## Overview

`rsahelpers` provides an end-to-end workflow for polynomial response surface
models estimated in Mplus. It can:

1. generate conventional measurement and structural syntax with tidySEM;
2. add the latent `XWITH` interactions required for quadratic terms;
3. generate response-surface and principal-axis constraints;
4. export an Mplus input file and its data;
5. run and read the model through MplusAutomation; and
6. extract and plot the fitted surface with `RSA_mplus()`.

The workflow supports two measurement strategies:

- **latent LMS**, with multiple indicators for each RSA variable; and
- **SI-LMS**, with one prepared score per construct and a fixed measurement
  error based on a supplied reliability estimate.

::: {.callout-note}
Mplus is needed only to estimate a generated input file. Model construction,
syntax inspection, and reading an existing `.out` file do not invoke the Mplus
executable.
:::

```{r}
#| label: setup
#| include: false

library(rsahelpers)

set.seed(2026)
n <- 300
x <- stats::rnorm(n)
y <- 0.35 * x + stats::rnorm(n)
z <- 0.30 * x + 0.20 * y - 0.10 * x^2 + 0.15 * x * y -
  0.08 * y^2 + stats::rnorm(n, sd = 0.7)

dat <- data.frame(
  x1 = x + stats::rnorm(n, sd = 0.4),
  x2 = x + stats::rnorm(n, sd = 0.4),
  x3 = x + stats::rnorm(n, sd = 0.4),
  y1 = y + stats::rnorm(n, sd = 0.4),
  y2 = y + stats::rnorm(n, sd = 0.4),
  y3 = y + stats::rnorm(n, sd = 0.4),
  z1 = z + stats::rnorm(n, sd = 0.4),
  z2 = z + stats::rnorm(n, sd = 0.4),
  z3 = z + stats::rnorm(n, sd = 0.4),
  age = sample(18:65, n, replace = TRUE)
)

analysis_variables <- c(
  "x1", "x2", "x3",
  "y1", "y2", "y3",
  "z1", "z2", "z3"
)
```

## Model and parameter conventions

The package uses the common RSA convention in which `Z` is the outcome and
`X` and `Y` are commensurate predictors:

$$
Z = b_0 + b_1X + b_2Y + b_3X^2 + b_4XY + b_5Y^2 + e.
$$

The generated latent interactions are named `XS`, `XY`, and `YS`. The five
structural paths are labeled `b1` through `b5`, which lets Mplus calculate the
surface parameters in `MODEL CONSTRAINT`.

| Mplus label | RSA parameter | Formula | Interpretation |
|---|---:|---|---|
| `cs` | $a_1$ | $b_1 + b_2$ | Slope along the line of congruence |
| `cc` | $a_2$ | $b_3 + b_4 + b_5$ | Curvature along the line of congruence |
| `is` | $a_3$ | $b_1 - b_2$ | Slope along the line of incongruence |
| `ic` | $a_4$ | $b_3 - b_4 + b_5$ | Curvature along the line of incongruence |
| `a5` | $a_5$ | $b_3 - b_5$ | Difference between predictor curvatures |

The default constraints also calculate the stationary point and the intercepts
and slopes of both principal axes: `x0`, `y0`, `p10`, `p11`, `p20`, and `p21`.

## Checking predictor coverage

Response surfaces should not be interpreted in combinations of X and Y that
are unsupported by the data. For scale scores derived from ordinal items,
`predictor_coverage()` uses common equal-width bins across the pooled X/Y range.
This preserves the geometry of `X = Y` without creating a cell for every
possible decimal score.

```{r}
#| label: fig-predictor-coverage
#| fig-cap: "Joint coverage of the two scale-score predictors. Red crosses mark bins containing fewer than five observations."
#| fig-alt: "Heatmap of joint X and Y scale-score coverage with sparse bins marked by red crosses."

scale_scores <- data.frame(
  X = rowMeans(dat[c("x1", "x2", "x3")]),
  Y = rowMeans(dat[c("y1", "y2", "y3")])
)
coverage <- predictor_coverage(
  scale_scores,
  X,
  Y,
  breaks = 10,
  discrepancy = c(0.5, 1)
)
coverage$summary
coverage$discrepancy
plot_predictor_coverage(coverage)
```

The summary uses neutral X/Y direction labels. Substantive labels such as
deficiency or excess depend on the constructs assigned to X and Y and should be
applied during interpretation rather than inferred by the function.

## Multi-indicator latent LMS

### List input

The most concise specification is a named list mapping each factor to its
indicators. `roles` defaults to `c(x = "X", y = "Y", outcome = "Z")`, so it
does not need to be supplied when those names are used.

```{r}
latent_model <- create_rsa_mplus_model(
  data = dat,
  measurement = list(
    X = c("x1", "x2", "x3"),
    Y = c("y1", "y2", "y3"),
    Z = c("z1", "z2", "z3")
  ),
  structural = list(Z = c("X", "Y")),
  usevariables = analysis_variables
)

latent_model
```

The core `Z ON X` and `Z ON Y` paths are added when they are absent, so the
`structural` argument can be omitted for the basic RSA model. A structural list
is useful when the outcome has additional predictors:

```{r}
latent_with_age <- create_rsa_mplus_model(
  data = dat,
  measurement = list(
    X = c("x1", "x2", "x3"),
    Y = c("y1", "y2", "y3"),
    Z = c("z1", "z2", "z3")
  ),
  structural = list(Z = c("X", "Y", "age")),
  usevariables = c(analysis_variables, "age")
)
```

### Character syntax

Measurement and structural models can instead be supplied as
lavaan/tidySEM-style character vectors. This form is useful for labeled
loadings, covariances, intercepts, and larger structural models.

```{r}
character_model <- create_rsa_mplus_model(
  data = dat,
  measurement = c(
    "X =~ x1 + x2 + x3",
    "Y =~ y1 + y2 + y3",
    "Z =~ z1 + z2 + z3"
  ),
  structural = c(
    "Z ~ X + Y + age",
    "X ~~ Y"
  ),
  usevariables = c(analysis_variables, "age")
)
```

### Inspecting generated syntax

The returned `rsa_mplus_workflow` stores both the tidySEM representation and
the assembled `mplusObject`.

```{r}
cat(latent_model$mplus$VARIABLE)
```

```{r}
model_lines <- strsplit(latent_model$mplus$MODEL, "\n", fixed = TRUE)[[1]]
cat(grep(" BY | ON |XWITH", model_lines, value = TRUE), sep = "\n")
```

```{r}
cat(latent_model$mplus$MODELCONSTRAINT)
```

Use `tidySEM::syntax(latent_model$tidysem)` when the underlying parameter table
is more useful than the rendered Mplus statements.

## Single-indicator LMS

SI-LMS uses one prepared score for each construct. If $r$ is the supplied
reliability and $s^2$ is the sample variance of the score, its residual variance
is fixed to

$$
\theta = (1-r)s^2.
$$

The score columns must be numeric, and reliabilities must be greater than zero
and no greater than one.

```{r}
scores <- data.frame(
  xmean = rowMeans(dat[c("x1", "x2", "x3")]),
  ymean = rowMeans(dat[c("y1", "y2", "y3")]),
  zmean = rowMeans(dat[c("z1", "z2", "z3")])
)

si_lms_model <- create_rsa_mplus_model(
  data = scores,
  measurement = list(
    X = "xmean",
    Y = "ymean",
    Z = "zmean"
  ),
  model_type = "si_lms",
  reliability = c(X = 0.82, Y = 0.79, Z = 0.88)
)

si_lines <- strsplit(si_lms_model$mplus$MODEL, "\n", fixed = TRUE)[[1]]
cat(grep(" BY |mean@", si_lines, value = TRUE), sep = "\n")
```

`rsahelpers` does not calculate the scores or their reliability estimates. This
keeps the scoring decisions explicit and allows alpha, omega, or another
appropriate reliability estimate to be supplied.

## Customizing Mplus blocks

`blocks` replaces generated defaults by Mplus section name. The default
`ANALYSIS` block uses `TYPE = RANDOM` and integration, and the default `OUTPUT`
block requests `STDYX`, confidence intervals, and `TECH1`.

```{r}
custom_model <- create_rsa_mplus_model(
  data = dat,
  measurement = list(
    X = c("x1", "x2", "x3"),
    Y = c("y1", "y2", "y3"),
    Z = c("z1", "z2", "z3")
  ),
  usevariables = c(analysis_variables, "age"),
  blocks = list(
    TITLE = "Latent RSA with four processors;",
    ANALYSIS = paste(
      "TYPE = RANDOM;",
      "ALGORITHM = INTEGRATION;",
      "PROCESSORS = 4;",
      sep = "\n"
    ),
    OUTPUT = "STDYX CINTERVAL TECH1 TECH4;"
  ),
  model_extra = "Z WITH age;",
  constraint_extra = "NEW(total_loc); total_loc = cs + cc;"
)
```

`TYPE = RANDOM` remains mandatory because the generated model contains
`XWITH`. `MODEL` and `MODEL CONSTRAINT` cannot be replaced through `blocks`;
use `model_extra` and `constraint_extra` to append statements without removing
the required RSA foundation.

To omit the stationary point and principal-axis formulas, request only the
surface constraints:

```{r}
surface_only <- create_rsa_mplus_model(
  data = dat,
  measurement = list(
    X = c("x1", "x2", "x3"),
    Y = c("y1", "y2", "y3"),
    Z = c("z1", "z2", "z3")
  ),
  constraints = "surface",
  usevariables = analysis_variables
)
```

This is useful when a principal-axis expression is undefined or unstable, such
as when $b_4$ or $4b_3b_5-b_4^2$ is near zero.

## Writing input and data files

The composable writer creates the `.inp` and `.dat` files without running
Mplus. Existing targets are protected unless `overwrite = TRUE`.

```{r}
#| eval: false

written <- write_rsa_mplus_model(
  latent_model,
  modelout = "models/latent-rsa.inp"
)

written$files
```

The convenience workflow performs model construction and file writing in one
call. Its default `run = FALSE` is deliberate.

```{r}
#| eval: false

written <- rsa_mplus_workflow(
  data = dat,
  measurement = list(
    X = c("x1", "x2", "x3"),
    Y = c("y1", "y2", "y3"),
    Z = c("z1", "z2", "z3")
  ),
  usevariables = analysis_variables,
  modelout = "models/latent-rsa.inp"
)
```

MplusAutomation writes every exported column under `NAMES`. rsahelpers also
writes an explicit `USEVARIABLES` statement containing the columns selected by
`usevariables`. This distinction matters: when `USEVARIABLES` is absent, Mplus
analyzes every variable under `NAMES`, including columns that were present only
for data management. The basic example therefore selects the nine indicators
and excludes `age`; the covariate examples include it intentionally.

By default `usevariables` contains all columns in `data`. Selected names must be
valid Mplus identifiers and unique in their first eight characters so parsed
output remains unambiguous. Other `VARIABLE` directives can be supplied through
`blocks$VARIABLE`, but `USEVARIABLES` itself is controlled by the dedicated
argument.

## Running and reading Mplus

Run a written workflow with `run_rsa_mplus_model()`:

```{r}
#| eval: false

fitted <- run_rsa_mplus_model(written)
```

The runner first checks an explicitly supplied executable and otherwise looks
for `mplus` or `Mplus` on the command path. An explicit path is useful when the
application is installed but not discoverable by the shell:

```{r}
#| eval: false

fitted <- run_rsa_mplus_model(
  written,
  Mplus_command = "/Applications/Mplus/mplus"
)
```

Set `run = TRUE` in `rsa_mplus_workflow()` to build, write, run, and parse in a
single call. Model files and Mplus output are retained if estimation fails.

An existing `.inp` or `.out` file can be read independently:

```{r}
#| eval: false

fitted <- read_rsa_mplus_model("models/latent-rsa.out")
```

This vignette reads a packaged output cache generated from the corrected model
above. Supplying it as `output` attaches the estimates to `latent_model` and
retains the role metadata. Thus the vignette evaluates result extraction and
plotting without requiring Mplus during package installation or rendering.

```{r}
cached_output <- system.file(
  "extdata",
  "mplus-response-surfaces.out",
  package = "rsahelpers"
)
if (!nzchar(cached_output)) {
  cache_candidates <- c(
    file.path("inst", "extdata", "mplus-response-surfaces.out"),
    file.path("..", "inst", "extdata", "mplus-response-surfaces.out")
  )
  cached_output <- cache_candidates[file.exists(cache_candidates)][[1]]
}

fitted <- read_rsa_mplus_model(latent_model, output = cached_output)
fitted
```

Standalone files do not contain the role metadata created by
`create_rsa_mplus_model()`. Consequently, plotting an arbitrary output requires
explicit variable labels.

## Extracting and plotting the response surface

`RSA_mplus()` accepts a fitted workflow directly. It infers `X`, `Y`, `Z`, and
the three interaction labels from workflow metadata.

```{r}
surface <- RSA_mplus(fitted, plot = FALSE)
surface$coefficients
surface$new_parameters[c("Label", "est", "se", "pval")]
```

## Classifying the surface

`classify_yao_ma()` reads the surface and principal-axis constraints directly
from the fitted Mplus model. Outcome valence is required because positively
valenced outcomes use the first principal axis (`P10`, `P11`), whereas
negatively valenced outcomes use the second (`P20`, `P21`).

```{r}
classification <- classify_yao_ma(fitted, valence = "positive")
classification
classification$decisions
```

By default, an interval excluding its null value indicates a difference and an
interval containing the null is treated as absence of a difference. For a
conservative equivalence analysis, supply separate margins for `CS`, `CC`,
`IS`, `IC`, the principal-axis intercept, and its slope. No margins are assumed
because their units depend on the model's working scale. Intervals that neither
fit fully inside nor fully outside the requested equivalence region produce an
indeterminate classification.

```{r}
#| eval: false

classify_yao_ma(
  fitted,
  valence = "positive",
  equivalence = c(
    CS = 0.10,
    CC = 0.10,
    IS = 0.10,
    IC = 0.10,
    axis_intercept = 0.15,
    axis_slope = 0.15
  )
)
```

Principal-axis constraints are nonlinear functions of the polynomial
coefficients. Their Mplus intervals use delta-method standard errors; unstable
or unavailable axes are therefore reported as a partial result rather than
being assigned a subtype.

### Publication-style coefficient table

The optional franzpak package can format both ordinary Mplus parameters and
`MODEL CONSTRAINT` results. Its function expects the parsed `mplus.model`,
which is stored in `fitted$results`.

```{r}
#| label: tbl-franzpak-mplus
#| tbl-cap: "Unstandardized Mplus estimates and model constraints."

franzpak::coef_table_mplus(fitted$results, constraints = TRUE)
```

Set `plot = TRUE` to pass the polynomial coefficients to `RSA::plotRSA()`.
Additional arguments are forwarded to that function.

```{r}
#| label: fig-mplus-rsa
#| fig-cap: "Latent response surface fitted in Mplus. Stars mark Mplus constraint p-values (p ≤ .05, .01, and .001)."
#| fig-alt: "Three-dimensional response surface with lines of congruence and incongruence and starred a1 through a5 parameters."
#| fig-width: 8
#| fig-height: 6

plotted <- RSA_mplus(
  fitted,
  plot = TRUE,
  xlab = "Actual (X)",
  ylab = "Desired (Y)",
  zlab = "Outcome (Z)"
)

plotted$plot
```

The surface itself is defined by `b0` and the five polynomial coefficients.
For the default three-dimensional plot, rsahelpers replaces RSA's calculated
parameter text with the `CS`, `CC`, `IS`, `IC`, and `A5` estimates from Mplus.
It adds `*`, `**`, and `***` for p-values no greater than .05, .01, and .001.
If these constraint estimates are unavailable, RSA's original unstarred text
is retained. `param = FALSE` hides the annotation, and contour plots do not
display a parameter block.

Generated latent and SI-LMS models fix the latent outcome mean to zero, so their
plotting intercept defaults to `b0 = 0`.

`coef_type` selects the Mplus coefficient table:

```{r}
RSA_mplus(fitted, coef_type = "un", plot = FALSE)
RSA_mplus(fitted, coef_type = "stdyx", plot = FALSE)
```

Unstandardized coefficients are the default and are generally the clearest
choice when the response surface is interpreted in the model's working scale.

## Choosing between latent LMS and SI-LMS

| Consideration | Latent LMS | SI-LMS |
|---|---|---|
| Indicators | Two or more per RSA factor | One prepared score per factor |
| Measurement error | Estimated from the measurement model | Fixed from reliability and score variance |
| Input required | Item-level columns | Scores plus reliability estimates |
| Flexibility | Factor-specific loadings and residuals | Simpler, fewer measurement parameters |
| Typical use | Primary latent-variable analysis | Too few indicators or convergence difficulties |

The two approaches estimate the same five-term polynomial structure. They
differ in how measurement error is represented, so switching between them is a
modeling decision rather than only a computational shortcut.

## Troubleshooting

- **Mplus is not found:** pass the executable through `Mplus_command` or add it
  to the system path.
- **The model has not been written:** call `write_rsa_mplus_model()` before the
  standalone runner, or use `rsa_mplus_workflow()`.
- **A role cannot be inferred:** workflow metadata is available only for models
  created by `create_rsa_mplus_model()`; supply all six labels for an arbitrary
  output.
- **Mplus names collide:** rename variables that share their first eight
  characters.
- **Principal-axis constraints fail:** use `constraints = "surface"` and inspect
  whether the fitted polynomial has singular principal-axis formulas.
- **SI-LMS construction fails:** confirm that each factor has exactly one
  numeric score and that reliability is named by the latent factors.
- **A model file already exists:** choose a new path or explicitly set
  `overwrite = TRUE`.

The generated object deliberately exposes `mplus`, `tidysem`, `files`, and
`results`, so the exact syntax, paths, parsed warnings, and estimates remain
available when diagnosing a model.
