---
title: "Spline Response Surfaces"
description: |
  Fit, compare, interpret, and plot one- and two-seam congruence spline models
  with rsahelpers.
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
vignette: >
  %\VignetteIndexEntry{Spline Response Surfaces}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Congruence spline regression represents a response surface with linear regions
joined at one or two estimated seam lines. Unlike polynomial response surface
analysis, the fitted surface can change slope abruptly where it crosses a seam.

This vignette uses the workshop data distributed with `rsahelpers` to show how
to:

1. prepare commensurate predictors;
2. fit the OLS models used for comparison and starting values;
3. estimate and interpret one- and two-seam spline surfaces;
4. compare constrained and unconstrained models;
5. visualize the fitted surface; and
6. obtain bootstrap confidence intervals.

## Formula convention

All spline entry points use the same formula structure:

```r
Z ~ X * Y
```

The response maps to Z, the first predictor to X, and the second predictor to
Y. The `*` declares the X and Y roles. It does not request the ordinary `X:Y`
interaction used by `lm()`.

> **Important:** Reversing the predictors reverses their X and Y roles. This
> matters because the estimated seam is written as a function of X,
> $Y = c_0 + c_1X$.

```{r}
#| label: setup
#| include: false

library(rsahelpers)

workshop <- haven::read_dta(
  system.file(
    "extdata",
    "spline.dta",
    package = "rsahelpers",
    mustWork = TRUE
  )
)
```

The main example models job satisfaction (`JOBSAT`) from the uncentered
authority variables `ATHW` (X) and `ATHH` (Y):

```{r}
authority_formula <- JOBSAT ~ ATHW * ATHH
```

The workshop data also contain precomputed variables with `*WC` and `*HC`
suffixes. Here we deliberately use the raw columns so the package applies its
default pooled centering and scaling, which preserve the meaning of the line
$X = Y$.

## Prepare the model data

`prepare_congruence_data()` applies complete-case selection, centering, and
scaling, then creates the hinge variables needed by the comparison models.

```{r}
authority_data <- prepare_congruence_data(
  authority_formula,
  data = workshop
)

head(authority_data[c("x", "y", "z", "abs_diff", "zd")])
```

The standardized output always uses `x`, `y`, and `z` internally. The source
variables and transformation constants remain attached as attributes.

```{r}
attr(authority_data, "variables")
c(
  x_center = attr(authority_data, "x_center"),
  y_center = attr(authority_data, "y_center"),
  scale = attr(authority_data, "scale")
)
```

## Fit comparison models

`fit_piecewise_congruence()` fits the absolute-difference, linear, one-break,
and constrained one-seam OLS models. Requesting two seams also adds a fixed
two-seam comparison.

```{r}
authority_ols <- fit_piecewise_congruence(
  authority_formula,
  data = workshop,
  n_seams = 2
)

tidy_piecewise_summary(authority_ols)
```

These models provide interpretable baselines and starting values for nonlinear
estimation. They are not substitutes for an estimated-seam spline.

## Estimate a one-seam surface

```{r}
authority_one <- fit_spline_congruence(
  authority_formula,
  data = workshop,
  n_seams = 1
)

summary(authority_one, warn_seam = FALSE)
coef(authority_one)
```

For a one-seam model, `c0` and `c1` define the seam
$Y = c_0 + c_1X$. The `b` coefficients define the plane on one side of the seam
and the change in slope after the seam is crossed.

`spline_surface_features()` translates these coefficients into region slopes,
seam features, symmetry expressions, and shifts along selected lines parallel
to the line of incongruence.

```{r}
spline_surface_features(authority_one)
```

`spline_tests()` supplies the corresponding scalar and joint Wald tests.

```{r}
authority_tests <- spline_tests(authority_one, warn_seam = FALSE)
authority_tests$scalar
authority_tests$joint
```

> **Warning:** Delta-method standard errors involving an estimated seam are
> approximate because the fitted surface is not differentiable at the seam.
> Use the bootstrap workflow below for final inference about seam features.

## Test a theoretically fixed seam

A seam fixed to the line $Y = X$ has `c0 = 0` and `c1 = 1`. The `fix` argument
removes those values from optimization, producing a constrained model that is
nested in the freely estimated one-seam spline.

```{r}
authority_fixed <- fit_spline_congruence(
  authority_formula,
  data = workshop,
  n_seams = 1,
  fix = c(c0 = 0, c1 = 1)
)

compare_spline_models(
  fixed_seam = authority_fixed,
  estimated_seam = authority_one
)
```

The comparison table reports fit statistics and the change in residual sum of
squares. Models supplied to `compare_spline_models()` must be ordered from
simpler to more complex.

## Estimate a two-seam surface

```{r}
authority_two <- fit_spline_congruence(
  authority_formula,
  data = workshop,
  n_seams = 2
)

summary(authority_two, warn_seam = FALSE)
spline_surface_features(authority_two)
anova(authority_one, authority_two)
```

Two-seam models can have local solutions or label-switched seams. By default,
`fit_spline_congruence()` tries several starting values and retains the
converged solution with the lowest residual sum of squares. A more complicated
model should not be retained merely because it contains another seam.

## Plot the fitted surface

A contour plot makes the seam location and supported regions easy to read.

```{r}
#| label: fig-authority-contour
#| fig-cap: "One-seam authority surface with the estimated seam and the line X = Y."
#| fig-alt: "Filled contour plot of predicted job satisfaction over X and Y, with an estimated seam and a dashed diagonal line."
#| fig-width: 7
#| fig-height: 6

plot_spline_contour(
  authority_one,
  xlab = "Authority X",
  ylab = "Authority Y",
  main = "Authority: one-seam spline"
)
```

The three-dimensional view emphasizes changes in the surface slope.

```{r}
#| label: fig-authority-surface
#| fig-cap: "Three-dimensional view of the one-seam authority surface."
#| fig-alt: "Three-dimensional tiled response surface for job satisfaction with an estimated seam line."
#| fig-width: 8
#| fig-height: 6

plot_spline_surface(
  authority_one,
  xlab = "Authority X",
  ylab = "Authority Y",
  zlab = "Job satisfaction",
  main = "Authority: one-seam spline"
)
```

## Bootstrap seam inference

Edwards and Parry recommend bootstrap intervals for the nonlinear seam
parameters and derived surface features. A final analysis should use a large
number of resamples; the code is not executed during vignette rendering.

```{r}
#| eval: false

set.seed(2026)
authority_boot <- bootstrap_spline(
  authority_one,
  R = 10000,
  type = "bca"
)
authority_boot
attr(authority_boot, "fail_rate")
```

Inspect the reported failure rate. A substantial rate indicates that bootstrap
samples frequently do not support a stable nonlinear solution.

## Repeat the workflow for another predictor pair

The variety variables use exactly the same Z ~ X * Y logic.

```{r}
variety_formula <- JOBSAT ~ VARW * VARH

variety_one <- fit_spline_congruence(
  variety_formula,
  data = workshop
)

summary(variety_one, warn_seam = FALSE)
spline_surface_features(variety_one)
```

Keeping the formula visible in each analysis makes the X/Y orientation
auditable while the remaining spline workflow stays unchanged.
