---
title: "Spline Response Surfaces"
description: |
  Fit, compare, interpret, and plot one- and two-seam congruence spline models
  with rsahelpers.
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
vignette: >
  %\VignetteIndexEntry{Spline Response Surfaces}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Congruence spline regression represents a response surface with linear regions
joined at one or two estimated seam lines. Unlike polynomial response surface
analysis, the fitted surface can change slope abruptly where it crosses a seam.

This vignette uses the workshop data distributed with `rsahelpers` to show how
to:

1. prepare commensurate predictors;
2. fit the OLS models used for comparison and starting values;
3. estimate and interpret one- and two-seam spline surfaces;
4. compare constrained and unconstrained models;
5. visualize the fitted surface; and
6. obtain bootstrap confidence intervals.

## Formula convention

All spline entry points use the same formula structure:

```r
Z ~ X * Y
```

The response maps to Z, the first predictor to X, and the second predictor to
Y. The `*` declares the X and Y roles. It does not request the ordinary `X:Y`
interaction used by `lm()`.

> **Important:** Reversing the predictors reverses their X and Y roles. This
> matters because the estimated seam is written as a function of X,
> $Y = c_0 + c_1X$.

```{r}
#| label: setup
#| include: false

library(rsahelpers)

workshop <- haven::read_dta(
  system.file(
    "extdata",
    "spline.dta",
    package = "rsahelpers",
    mustWork = TRUE
  )
)
```

The main example models job satisfaction (`JOBSAT`) from the uncentered
authority variables `ATHW` (X) and `ATHH` (Y):

```{r}
authority_formula <- JOBSAT ~ ATHW * ATHH
```

The workshop data also contain precomputed variables with `*WC` and `*HC`
suffixes. Here we deliberately use the raw columns so the package applies its
default pooled centering and scaling, which preserve the meaning of the line
$X = Y$.

## Prepare the model data

`prepare_congruence_data()` applies complete-case selection, centering, and
scaling, then creates the hinge variables needed by the comparison models.

```{r}
authority_data <- prepare_congruence_data(
  authority_formula,
  data = workshop
)

head(authority_data[c("x", "y", "z", "abs_diff", "zd")])
```

The standardized output always uses `x`, `y`, and `z` internally. The source
variables and transformation constants remain attached as attributes.

```{r}
attr(authority_data, "variables")
c(
  x_center = attr(authority_data, "x_center"),
  y_center = attr(authority_data, "y_center"),
  scale = attr(authority_data, "scale")
)
```

## Fit comparison models

`fit_piecewise_congruence()` fits the absolute-difference, linear, one-break,
and constrained one-seam OLS models. Requesting two seams also adds a fixed
two-seam comparison.

```{r}
authority_ols <- fit_piecewise_congruence(
  authority_formula,
  data = workshop,
  n_seams = 2
)

tidy_piecewise_summary(authority_ols)
```

These models provide interpretable baselines and starting values for nonlinear
estimation. They are not substitutes for an estimated-seam spline.

## Estimate a one-seam surface

```{r}
authority_one <- fit_spline_congruence(
  authority_formula,
  data = workshop,
  n_seams = 1
)

summary(authority_one, warn_seam = FALSE)
coef(authority_one)
```

For a one-seam model, `c0` and `c1` define the seam
$Y = c_0 + c_1X$. The `b` coefficients define the plane on one side of the seam
and the change in slope after the seam is crossed.

`spline_surface_features()` translates these coefficients into region slopes,
seam features, symmetry expressions, and shifts along selected lines parallel
to the line of incongruence.

```{r}
spline_surface_features(authority_one)
```

`spline_tests()` supplies scalar and joint Wald tests for properties of a
supported one-seam surface. It deliberately does not use a Wald test to decide
whether a seam exists: under the no-seam null, its location is unidentified.

```{r}
authority_tests <- spline_tests(authority_one, warn_seam = FALSE)
authority_tests$scalar
authority_tests$joint
```

> **Warning:** Delta-method standard errors involving an estimated seam are
> approximate because the fitted surface is not differentiable at the seam.
> Use the bootstrap workflow below for final inference about seam features.

## Test a theoretically fixed seam

A seam fixed to the line $Y = X$ has `c0 = 0` and `c1 = 1`. The `fix` argument
removes those values from optimization, producing a constrained model that is
nested in the freely estimated one-seam spline.

```{r}
authority_fixed <- fit_spline_congruence(
  authority_formula,
  data = workshop,
  n_seams = 1,
  fix = c(c0 = 0, c1 = 1)
)

compare_spline_models(
  fixed_seam = authority_fixed,
  estimated_seam = authority_one
)
```

The comparison table reports fit statistics and the change in residual sum of
squares. This nested fixed-versus-free comparison is interpretable only after
the data support a seam. Its ordinary F reference is a large-sample
approximation for a nonlinear kink model; the selection function below uses a
null bootstrap for this restriction as well.

## Select the continuous surface

`select_spline_congruence()` applies the complete decision sequence. It first
uses a null residual-bootstrap likelihood-ratio test for a seam, checks that
both arms are populated, and only then tests whether the supported seam can be
fixed to the line of congruence.

### What enters the null residual-bootstrap LRT?

The first structural comparison is a linear plane against a freely located
one-seam spline. On the transformed working scale used by the fitted models,
these are

$$
H_0: z_i = b_0 + b_1x_i + b_2y_i + e_i
$$

and

$$
H_1: z_i = b_0 + b_1x_i + b_2y_i
  + b_3h_i(c_0,c_1) + e_i,
$$

where

$$
h_i(c_0,c_1) = (y_i-c_0-c_1x_i)
  \mathbb{1}(y_i<c_0+c_1x_i).
$$

The null is therefore “no change in slope,” $b_3=0$. Under that null, every
value of $(c_0,c_1)$ gives the same linear plane: the seam location is not
identified. This is the nuisance-parameter-under-the-alternative problem that
invalidates the usual chi-squared likelihood-ratio, extra-sum-of-squares F, and
joint Wald reference distributions. It is a standard issue in unknown-threshold
and regression-kink testing; see [Davies
(1977)](https://doi.org/10.1093/biomet/64.2.247) and [Hansen
(2017)](https://users.ssc.wisc.edu/~behansen/papers/cthresh.html).

For the observed data, the function computes the Gaussian profile
likelihood-ratio statistic

$$
T_{obs}=2(\ell_1-\ell_0)
       =n\log\left(\frac{RSS_0}{RSS_1}\right),
$$

using the same complete cases and the same transformed X and Y in both fits.
Its null distribution is then approximated as follows:

1. Fit the null model and retain its fitted values and centered residuals.
2. Generate a null outcome
   $z_i^*=\widehat z_{0i}+e_i^*$ by sampling the centered null residuals with
   replacement. X and Y remain fixed.
3. Refit both the null plane and the free one-seam spline to $z^*$. The spline
   refit repeats the multistart search, so its search over seam locations is
   represented in the null distribution.
4. Recompute $T^*=n\log(RSS_0^*/RSS_1^*)$ and repeat this `R` times.
5. Report
   $(1 + \#\{T^*\ge T_{obs}\})/(1+B)$, where $B$ is the number of successful
   paired refits. The added ones avoid a zero Monte Carlo p-value.

`successful` and `fail_rate` in the test output make the nonlinear-refit
quality visible. If the failed proportion exceeds `max_fail`, selection is
marked inconclusive instead of treating the available replicates as a reliable
test. Even below that cutoff, dropping failed refits can bias the p-value if
failure is related to unusually easy or difficult null samples, so a nonzero
failure rate still deserves investigation.

This is specifically a **fixed-X/Y, residual bootstrap**. Resampling residuals
assumes independent, similarly distributed, approximately homoskedastic
errors. It does not make the test robust to heteroskedasticity, clusters, or
serial dependence; those settings need an appropriate wild, cluster, or block
null bootstrap. The default `R = 1999` also leaves visible Monte Carlo
uncertainty near a decision boundary. Use substantially more replications for
a final analysis and check whether the conclusion is stable.

### The complete selection sequence

The procedure uses an ordered, theory-driven comparison rather than asking all
candidate models to compete at once:

1. Test the linear plane against the free one-seam spline with the null
   residual-bootstrap LRT.
2. If the test does not reject, retain the linear plane. If the bootstrap is
   unreliable, report `inconclusive_seam_test` while conservatively retaining
   the linear fit.
3. If a seam is supported, require at least `min_arm_n` observations and
   `min_arm_prop` of the sample on each side of the estimated seam. These are
   identification diagnostics, not another hypothesis test. Failure produces
   `weakly_identified_seam` and the linear fallback.
4. Only after those gates, compare the spline fixed at $Y=X$ with the freely
   located spline using a second null residual bootstrap. Not rejecting the
   restriction selects the more parsimonious fixed-LOC surface; it does not
   prove that the population seam is exactly $Y=X$.
5. If that second bootstrap is inconclusive, retain the free one-seam model and
   report `free_seam_location_inconclusive`.

For the second test, the null fit includes the hinge contribution $b_3$ but
fixes $(c_0,c_1)=(0,1)$. Its fitted values and centered residuals generate the
null outcomes, and every replicate refits both that fixed-LOC spline and the
free-LOC spline. The observed and replicate statistics use the same RSS-ratio
formula as the seam-existence test.

Both decisions use the unadjusted `alpha` supplied by the user. The sequence is
not a single omnibus test and does not provide a family-wise error guarantee.
The absolute-difference and discontinuous piecewise fits remain benchmarks.
The optional two-seam fit remains a sensitivity analysis and cannot become the
selected focal model.

```{r}
# Use a larger R for final inference.
authority_selection <- select_spline_congruence(
  authority_formula,
  data = workshop,
  R = 99,
  seed = 20260722
)

authority_selection
authority_selection$tests
authority_selection$diagnostics
```

### How this compares with common alternatives

For models fit to the same outcome observations, define the partial R-squared
for the larger model as

$$
R^2_{partial}=\frac{RSS_0-RSS_1}{RSS_0}.
$$

Then $T_{obs}=-n\log(1-R^2_{partial})$. Thus, the LRT and partial R-squared rank
the observed improvement identically. The bootstrap is not finding a different
improvement; it supplies the non-standard null reference distribution.

| Method | What it answers here | Appropriate role |
|---|---|---|
| Delta or partial R-squared | How much in-sample variation/RSS the spline adds | Report as effect size. It has no valid generic cutoff or ordinary F calibration when the seam is unidentified. |
| Extra-sum-of-squares F test | Whether added terms improve a regular nested model | Do not use its usual F distribution for linear versus free-seam or one- versus two-seam tests. A specially bootstrapped F statistic would be a reasonable alternative calibration. |
| AIC | Which candidate has lower estimated expected information loss | Useful as descriptive/predictive sensitivity evidence, not as a level-$\alpha$ seam test. Standard parameter-count penalties are only approximate for this non-regular alternative. |
| AICc | AIC with a larger small-sample penalty | Potentially useful when $n$ is not large relative to the number of parameters, but the usual correction assumes regular model dimensions and does not repair unidentified seam parameters. |
| BIC | Which candidate wins under a stronger, sample-size-dependent penalty | A conservative descriptive sensitivity check. Standard BIC's usual marginal-likelihood interpretation also relies on regular identification; singular-model variants require additional machinery not implemented here. |
| Wald test | Whether a coefficient or a regular set of restrictions differs from its null | A naive test of $b_3=0$, or of $(b_3,c_0,c_1)$ jointly, is invalid for seam existence because $(c_0,c_1)$ disappear under the null. Conditional seam-location Wald tests are only approximate at the kink; prefer the bootstrap. |
| Cross-validation | Whether the spline predicts held-out outcomes better | Useful if prediction is the goal, but it does not test the scientific null that a seam exists or lies on $Y=X$. |

For orientation, the free one-seam model adds three estimated parameters over
the plane: $b_3$, $c_0$, and $c_1$. Ordinary AIC prefers it when
$T_{obs}>2\times3=6$, whereas BIC prefers it when
$T_{obs}>3\log(n)$. Those are penalized-selection rules, not p-values, and
their usual derivations treat all three parameters as regularly identified.
That assumption is precisely what fails under the no-seam null. More general
criteria for singular models exist, but require a different fitting framework;
see [Watanabe (2013)](https://www.jmlr.org/papers/v14/watanabe13a.html).

The ordinary fit indices can still be inspected side by side as sensitivity
evidence:

```{r}
selection_candidates <- authority_selection$fits[
  c("linear", "fixed_LOC_spline", "one_seam_spline")
]

candidate_fit <- function(model) {
  r2 <- if (inherits(model, "congruence_spline")) {
    model$r.squared
  } else {
    summary(model)$r.squared
  }
  c(R2 = r2, AIC = AIC(model), BIC = BIC(model))
}

t(vapply(selection_candidates, candidate_fit, numeric(3)))
```

Use these criteria to describe concordance or tension with the bootstrap
decision, not to replace its p-value mechanically. Likewise,
`compare_spline_models()` can summarize delta R-squared for genuinely nested
fits, but its ordinary F p-value is not the structural seam-existence test.

## Estimate a two-seam surface

```{r}
authority_two <- fit_spline_congruence(
  authority_formula,
  data = workshop,
  n_seams = 2
)

summary(authority_two, warn_seam = FALSE)
spline_surface_features(authority_two)
```

Two-seam models can have local solutions or label-switched seams. By default,
`fit_spline_congruence()` tries several starting values and retains the
converged solution with the lowest residual sum of squares. A conventional F
test is not valid for one versus two seams because the second seam's location
is unidentified under the one-seam null; here the two-seam surface is
descriptive sensitivity evidence only.

## Plot the fitted surface

A contour plot makes the seam location and supported regions easy to read.

```{r}
#| label: fig-authority-contour
#| fig-cap: "One-seam authority surface with the estimated seam and the line X = Y."
#| fig-alt: "Filled contour plot of predicted job satisfaction over X and Y, with an estimated seam and a dashed diagonal line."
#| fig-width: 7
#| fig-height: 6

plot_spline_contour(
  authority_one,
  xlab = "Authority X",
  ylab = "Authority Y",
  main = "Authority: one-seam spline"
)
```

The three-dimensional view emphasizes changes in the surface slope.

```{r}
#| label: fig-authority-surface
#| fig-cap: "Three-dimensional view of the one-seam authority surface."
#| fig-alt: "Three-dimensional tiled response surface for job satisfaction with an estimated seam line."
#| fig-width: 8
#| fig-height: 6

plot_spline_surface(
  authority_one,
  xlab = "Authority X",
  ylab = "Authority Y",
  zlab = "Job satisfaction",
  main = "Authority: one-seam spline"
)
```

## Bootstrap seam inference

Edwards and Parry recommend bootstrap intervals for the nonlinear seam
parameters and derived surface features. A final analysis should use a large
number of resamples; the code is not executed during vignette rendering.

```{r}
#| eval: false

set.seed(2026)
authority_boot <- bootstrap_spline(
  authority_one,
  R = 10000,
  type = "bca"
)
authority_boot
attr(authority_boot, "fail_rate")
```

Inspect the reported failure rate. A substantial rate indicates that bootstrap
samples frequently do not support a stable nonlinear solution.

## Repeat the workflow for another predictor pair

The variety variables use exactly the same Z ~ X * Y logic.

```{r}
variety_formula <- JOBSAT ~ VARW * VARH

variety_one <- fit_spline_congruence(
  variety_formula,
  data = workshop
)

summary(variety_one, warn_seam = FALSE)
spline_surface_features(variety_one)
```

Keeping the formula visible in each analysis makes the X/Y orientation
auditable while the remaining spline workflow stays unchanged.
